Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-21 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
minor updates to section 1 and 2 - basically clean up after chat and Shannon's edits

The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process, as has been seen in the current COVID-19 \shannon{\sout{epi}}\shannon{pan}demic \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}. \footnote{\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}\shannon{Seems fine to me now}}

With \shannon{\sout{the}} many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} \shannon{available}, this field has a \shannon{\sout{lot of}} \shannon{number of} devices to \shannon{\sout{aid}} \shannon{help}  an individual practitioner decide the correct approach. For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.

@@

One of \pkg{EpiCompare}’s main contribution to comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments. Epidemics, despite being defined as a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales \shannon{in order}\footnote{\shannon{wanted a clearer signal that this is a bridge to a new thought}}to understand the processes at play. With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible. Compared to time-dependent comparison tools for state-space modeling,  time-invariant analysis can make it easier to compare state-space epidemic representations in a more global, holistic fashion. Many time-dependent comparison tools only examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion. These time-dependent approaches can reduce the amount of connections that can be seen and insights that can be drawn}, similar to examining projections of a multidimensional distribution onto a single axis, one at a time. Tools in \pkg{EpiCompare} \shannon{\sout{give the user the ability to extend their} \shannon{extend the user}} toolkit to evaluate epidemics within a time-invariant lens. At the same time, the goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a  demonstration of the tools through a detailed example\shannon{\sout{ of a full data analysis pipeline.}} \shannon{from start to finish of the data analysis pipeline}\footnote{\shannon{maybe clearer? up to you}}

@@

\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis. In this section we highlight some weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take. 
To accomplish this goal we  first demonstrate the inability of time-dependent tools to \ben{adequately} quantify\footnote{\ben{[Ben says: original suggestion was "simply capture", and original phrase was "well quantify".]}} a classic epidemic parameter, the reproduction number $R_0$. 
Then we motivate new time-invariant approaches for more complex situations where $R_0$ does not capture the complexities of outbreaks. This leads to the final subsection that defines ways to view epidemics in a time-invariant lens and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.

@@

Epidemiological research is often interested in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected. One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.  This \ben{numerical summary} \sout{intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}. \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic. This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, intrinsic properties of an epidemic.

@@

Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space \sout{(i.e. through a time-invariant lens)}\footnote{\ben{[Ben says: the next sentence mentions time-invariance - where you thinking something more global?]}} allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted. In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states. These tools leverage the idea that an epidemic can be viewed as a trajectory and that many properties of the epidemic are well captured when we do so. This approach is useful when the epidemic of interest has only gone through a single realization of its outbreak (before the population of individuals become susceptible again).

@@

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
\]
where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$ (where $|\psi_i|$ is the length of $\psi_i$). This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness estimators can be useful when comparing the true epidemic to a set of simulated epidemics, and practitioners can interpret the epidemic's extremeness score relative to the extremeness scores of the simulations very easily.

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process, as has been seen in the current COVID-19 pandemic \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.

With many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} available, this field has a number of devices to \shannon{\sout{aid}} \shannon{help}\footnote{\ben{[Ben says: I'm not sure this an improvement]}}  an individual practitioner decide the correct approach. For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.

@@

One of \pkg{EpiCompare}’s main contribution to comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments. Epidemics, despite being defined as a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales in order to understand the processes at play. With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible. Compared to time-dependent comparison tools for state-space modeling,  time-invariant analysis can make it easier to compare state-space epidemic representations in a more global, holistic fashion. Many time-dependent comparison tools only examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion. These time-dependent approaches can reduce the amount of connections that can be seen and insights that can be drawn}, similar to examining projections of a multidimensional distribution onto a single axis, one at a time. Tools in \pkg{EpiCompare} extend the user toolkit to evaluate epidemics within a time-invariant lens. At the same time, the goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a  demonstration of the tools through a detailed example\shannon{\sout{ of a full data analysis pipeline.}} \shannon{from start to finish of the data analysis pipeline}\footnote{\shannon{maybe clearer? up to you} \ben{The current suggestion isn't clearer - feel free to try again.}}

@@

\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis. In this section we highlight some weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take. 
To accomplish this goal we  first demonstrate the inability of time-dependent tools to adequately quantify a classic epidemic parameter, the reproduction number $R_0$. 
Then we motivate new time-invariant approaches for more complex situations where $R_0$ does not capture the complexities of outbreaks. This leads to the final subsection that defines ways to view epidemics in a time-invariant lens and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.

@@

Epidemiological research is often interested in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected. One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.  This numerical summary is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}. \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic. This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, intrinsic properties of an epidemic.

@@

Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted. In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states. These tools leverage the idea that an epidemic can be viewed as a trajectory and that many properties of the epidemic are well captured when we do so. This approach is useful when the epidemic of interest has only gone through a single realization of its outbreak (before the population of individuals become susceptible again).

@@

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
\]
where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$ (with $|\psi_i|$ is the length of $\psi_i$). This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness estimators can be useful when comparing the true epidemic to a set of simulated epidemics, and practitioners can interpret the epidemic's extremeness score relative to the extremeness scores of the simulations very easily.

### 2021-05-21 jss/EpiCompare-paper/EpiCompare-paper2.tex
minor updates to section 1 and 2 - basically clean up after chat and Shannon's edits

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process, as has been seen in the current COVID-19
\textcolor{orange}{\sout{epi}}\textcolor{orange}{pan}demic
\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}\textcolor{orange}{Seems fine to me now}}

With \textcolor{orange}{\sout{the}} many visualization and exploratory
tools, models and modeling paradigms, and reviews and comparisons in the
literature and through the MIDAS (Models of Infectious Disease Agent
Study) network \citep{midasnetwork2021} \textcolor{orange}{available},
this field has a \textcolor{orange}{\sout{lot of}}
\textcolor{orange}{number of} devices to \textcolor{orange}{\sout{aid}}
\textcolor{orange}{help} an individual practitioner decide the correct
approach. For example, \proglang{R} packages such as \pkg{surveillance},
\pkg{EpiModel}, and \pkg{pomp} have all made significant steps in
standardizing the flow of the data analysis pipeline for epidemic
modeling through digitizing data sets, making accessible statistical
models, and providing a plethora of educational material for both coding
novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.

@@

One of \pkg{EpiCompare}'s main contribution to comparison and assessment
of epidemics is through tools that provide \textit{time-invariant}
assessments. Epidemics, despite being defined as a process that evolves
over time, often need to be compared in a way not constrained to initial
times or time scales
\textcolor{orange}{in order}\footnote{\textcolor{orange}{wanted a clearer signal that this is a bridge to a new thought}}to
understand the processes at play. With time-invariant analysis,
comparing decades-long outbreaks of HIV in the US to a 10 day outbreak
of norovirus on a cruise ship is possible. Compared to time-dependent
comparison tools for state-space modeling, time-invariant analysis can
make it easier to compare state-space epidemic representations in a more
global, holistic fashion. Many time-dependent comparison tools only
examine the proportion of individuals in each state (at a given time) in
a piece-wise / marginal fashion. These time-dependent approaches can
reduce the amount of connections that can be seen and insights that can
be drawn\}, similar to examining projections of a multidimensional
distribution onto a single axis, one at a time. Tools in
\pkg{EpiCompare}
\textcolor{orange}{\sout{give the user the ability to extend their} \textcolor{orange}{extend the user}}
toolkit to evaluate epidemics within a time-invariant lens. At the same
time, the goal of \pkg{EpiCompare} is not to supplant existing
infectious disease modeling tools and software but, rather, is a
concerted effort to create standard and fair comparisons among models
developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a demonstration of the tools through a
detailed
example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer? up to you}}

@@

\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way - approaches that remove some or all of the
impact of start/end times and recording time scales when preforming the
analysis. In this section we highlight some weaknesses of
time-\textit{de}pendent analysis and define the mathematical
underpinning of the time-invariant approach we take. To accomplish this
goal we first demonstrate the inability of time-dependent tools to
\textcolor{violet}{adequately}
quantify\footnote{\textcolor{violet}{[Ben says: original suggestion was "simply capture", and original phrase was "well quantify".]}}
a classic epidemic parameter, the reproduction number \(R_0\). Then we
motivate new time-invariant approaches for more complex situations where
\(R_0\) does not capture the complexities of outbreaks. This leads to
the final subsection that defines ways to view epidemics in a
time-invariant lens and discusses natural properties of these
representations that allow for clearer ways to compare models and
epidemics.

@@

Epidemiological research is often interested in understanding intrinsic
properties of the epidemic, independent from when the epidemic occurred
or how frequently data was collected. One of the most famous numerical
summaries of an epidemic is the reproduction number - \(R_0\). This
\textcolor{violet}{numerical summary} \sout{intrinsic property} is a
time-invariant value that is defined as the expected number of
infections caused by a single infect or who is added to a completely
susceptible population \citep{anderson1992}. \citet{Gallagher2020}
showed that the estimation of \(R_0\) can be sensitive to time-dependent
parameters like estimation of the beginning and end of an epidemic. This
time-dependent sensitivity in estimating \(R_0\) reflects a general
problem in epidemiology surrounding how to transform time-dependent
information into insights into desirable, intrinsic properties of an
epidemic.

@@

Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space
\sout{(i.e. through a time-invariant lens)}\footnote{\textcolor{violet}{[Ben says: the next sentence mentions time-invariance - where you thinking something more global?]}}
allows us to better compare the overall structure of the epidemic and
see how the population is directly impacted. In this section we present
time-invariant tools that can be applied to complex epidemics where the
epidemic's generative process is unknown and can have more than three
states. These tools leverage the idea that an epidemic can be viewed as
a trajectory and that many properties of the epidemic are well captured
when we do so. This approach is useful when the epidemic of interest has
only gone through a single realization of its outbreak (before the
population of individuals become susceptible again).

@@

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (where
\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
time-invariant, and can be plugged into multiple distance-based
assessment tools to examine the overall ``extremeness'\,' of points,
including pseudo-density estimators and depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estimators can be useful when comparing the true epidemic to a set of
simulated epidemics, and practitioners can interpret the epidemic's
extremeness score relative to the extremeness scores of the simulations
very easily.

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process, as has been seen in the current COVID-19 pandemic
\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.

With many visualization and exploratory tools, models and modeling
paradigms, and reviews and comparisons in the literature and through the
MIDAS (Models of Infectious Disease Agent Study) network
\citep{midasnetwork2021} available, this field has a number of devices
to \textcolor{orange}{\sout{aid}}
\textcolor{orange}{help}\footnote{\textcolor{violet}{[Ben says: I'm not sure this an improvement]}}
an individual practitioner decide the correct approach. For example,
\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
\pkg{pomp} have all made significant steps in standardizing the flow of
the data analysis pipeline for epidemic modeling through digitizing data
sets, making accessible statistical models, and providing a plethora of
educational material for both coding novices and experts alike
\citep{surveillance2017,Jenness2018,King2016}.

@@

One of \pkg{EpiCompare}'s main contribution to comparison and assessment
of epidemics is through tools that provide \textit{time-invariant}
assessments. Epidemics, despite being defined as a process that evolves
over time, often need to be compared in a way not constrained to initial
times or time scales in order to understand the processes at play. With
time-invariant analysis, comparing decades-long outbreaks of HIV in the
US to a 10 day outbreak of norovirus on a cruise ship is possible.
Compared to time-dependent comparison tools for state-space modeling,
time-invariant analysis can make it easier to compare state-space
epidemic representations in a more global, holistic fashion. Many
time-dependent comparison tools only examine the proportion of
individuals in each state (at a given time) in a piece-wise / marginal
fashion. These time-dependent approaches can reduce the amount of
connections that can be seen and insights that can be drawn\}, similar
to examining projections of a multidimensional distribution onto a
single axis, one at a time. Tools in \pkg{EpiCompare} extend the user
toolkit to evaluate epidemics within a time-invariant lens. At the same
time, the goal of \pkg{EpiCompare} is not to supplant existing
infectious disease modeling tools and software but, rather, is a
concerted effort to create standard and fair comparisons among models
developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a demonstration of the tools through a
detailed
example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer? up to you} \textcolor{violet}{The current suggestion isn't clearer - feel free to try again.}}

@@

\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way - approaches that remove some or all of the
impact of start/end times and recording time scales when preforming the
analysis. In this section we highlight some weaknesses of
time-\textit{de}pendent analysis and define the mathematical
underpinning of the time-invariant approach we take. To accomplish this
goal we first demonstrate the inability of time-dependent tools to
adequately quantify a classic epidemic parameter, the reproduction
number \(R_0\). Then we motivate new time-invariant approaches for more
complex situations where \(R_0\) does not capture the complexities of
outbreaks. This leads to the final subsection that defines ways to view
epidemics in a time-invariant lens and discusses natural properties of
these representations that allow for clearer ways to compare models and
epidemics.

@@

Epidemiological research is often interested in understanding intrinsic
properties of the epidemic, independent from when the epidemic occurred
or how frequently data was collected. One of the most famous numerical
summaries of an epidemic is the reproduction number - \(R_0\). This
numerical summary is a time-invariant value that is defined as the
expected number of infections caused by a single infect or who is added
to a completely susceptible population \citep{anderson1992}.
\citet{Gallagher2020} showed that the estimation of \(R_0\) can be
sensitive to time-dependent parameters like estimation of the beginning
and end of an epidemic. This time-dependent sensitivity in estimating
\(R_0\) reflects a general problem in epidemiology surrounding how to
transform time-dependent information into insights into desirable,
intrinsic properties of an epidemic.

@@

Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space allows us
to better compare the overall structure of the epidemic and see how the
population is directly impacted. In this section we present
time-invariant tools that can be applied to complex epidemics where the
epidemic's generative process is unknown and can have more than three
states. These tools leverage the idea that an epidemic can be viewed as
a trajectory and that many properties of the epidemic are well captured
when we do so. This approach is useful when the epidemic of interest has
only gone through a single realization of its outbreak (before the
population of individuals become susceptible again).

@@

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (with
\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
time-invariant, and can be plugged into multiple distance-based
assessment tools to examine the overall ``extremeness'' of points,
including pseudo-density estimators and depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estimators can be useful when comparing the true epidemic to a set of
simulated epidemics, and practitioners can interpret the epidemic's
extremeness score relative to the extremeness scores of the simulations
very easily.
